Regional Base-Flow Index in Arid Landscapes Using Machine Learning and Instrumented Records

### 2025-01-25 MANUSCRIPT.qmd
starting submission docs

key-points:
  - Approximately 32% of Arizona's streamflow originates from groundwater discharge, with significant spatial and climatic variability. 
  - A machine learning model (XGBoost) successfully predicted long-term BFI in ungauged basins, addressing gaps in Arizonas sparse streamgage network. 
  - Base-flow trends align most closely with precipitation changes, emphasizing the critical role of climate in shaping groundwater-surface water dynamics in dryland regions.

@@

This study develops a technique for estimating BFI in ungauged basins across Arizona and evaluates the state’s long-term BFI. Regional trends in base flow and BFI at instrumented sites are also analyzed, with these trends being linked to coincident trends in precipitation, evapotranspiration (ET~O~), and temperature over the same periods. Using a machine learning model trained on the hydrogeologic characteristics of surface water basins, we estimate the mean BFI for ungauged basins from 1991 to 2020. This approach helps address the spatial gaps in the streamgage network, which is relatively sparse across the state. The results offer novel insights into low-flow processes in both gauged and ungauged basins, enhancing our understanding of climate controls on consistent flows in Arizona. This study contributes to a more comprehensive view of hydrological dynamics in the context of arid and semi-arid landscapes.

@@

Annual reference evapotranspiration (ET~O~) data came from TerraClimate, a 4-km grid climatalogical data set [@abatzoglou2018]. TerraClimate uses a Penman-Monteith approach to generate a reference evapotranspiration. The ET~O~ values were calculated assuming a reference grass surface across the landscape with unlimited water. In the drylands of the southwestern US, ET~O~ typically exceeds precipitation annually [@zomer2022].

@@

![Linear relationship between observed BFI and predicted BFI. The solid line is the 1:1 line, the dashed line is regressed to the data.](images/actual-predicted.png){#fig-actual_predicted}

@@

: Performance of model predictions for BFI for all sites split by various classifications. n is number of observatios, R^2^ is the coefficient of determination of a linear regression, MSE is mean-squared-error, RMSE is root-mean-squared-error, MAE is mean-absolute-error, NSE is Nash-Sucliffe efficiency, and pbias is percent bias. {#tbl-performance}

@@

Given the variations in the period of record across the instrumented network (see SUPPLEMENTAL - Period of Record figure), we analyzed trends in base flow and BFI in relation to coincident trends in climate variables. Trends are classified as coincident when the direction of the climate variable trend (positive or negative) aligns with the trend observed in base flow or BFI (@tbl-coincident-trends). This analysis includes both significant and non-significant trends, which is appropriate where the influence of complex, interconnected processes may not always manifest as statistically significant patterns over limited observational periods [@ficklin2016].

@@

However, the utility of the streamgage networks for base-flow analyses in Arizona is limited by their design and focus. There has been an increase in the number of streamgages in the region to meet regulatory imperatives, such as the Clean Water Act. Even so, many of these gages are not suited for base-flow studies because they emphasize peak flow monitoring and lack the ability to accurately measure low-flow dynamics [@flood-control2020]. New, non-USGS streamgages are typically installed by flood control districts (e.g. the ALERT system) and are designed for tracking flood flows rather than base flow. New USGS streamgages have been added over the past decade to address instream flow rights and to improve the density of monitoring in the future.

>>>>

key-points:
  - Approximately 32% of Arizona's streamflow originates from groundwater discharge, with significant spatial and climatic variability. 
  - A machine learning model (XGBoost) successfully predicted long-term BFI in ungauged basins, addressing gaps in a sparse streamgage network. 
  - Base-flow trends align most closely with precipitation changes, emphasizing the role of climate in shaping groundwater-surface water dynamics in dryland regions.

@@

This study develops a technique for estimating BFI in ungauged basins across Arizona and evaluates the state’s long-term BFI. Regional trends in base flow and BFI at instrumented sites are also analyzed, with these trends being linked to coincident trends in precipitation, reference evapotranspiration (ET~O~), and temperature over the same periods. Using a machine learning model trained on the hydrogeologic characteristics of surface water basins, we estimate the mean BFI for ungauged basins from 1991 to 2020. This approach helps address the spatial gaps in the streamgage network, which is relatively sparse across the state. The results offer novel insights into low-flow processes in both gauged and ungauged basins, enhancing our understanding of climate controls on consistent flows in Arizona. This study contributes to a more comprehensive view of hydrological dynamics in the context of arid and semi-arid landscapes.

@@

Annual reference evapotranspiration (ET~O~) data came from TerraClimate, a 4-km grid climatological data set [@abatzoglou2018]. TerraClimate uses a Penman-Monteith approach to generate a reference evapotranspiration. The ET~O~ values were calculated assuming a reference grass surface across the landscape with unlimited water. In the drylands of the southwestern US, ET~O~ typically exceeds precipitation annually [@zomer2022].

@@

![Linear relationship between observed BFI and predicted BFI. The solid line is the 1:1 line, the dashed, green line is regressed to the data.](images/actual-predicted.png){#fig-actual_predicted}

@@

: Performance of model predictions for BFI for all sites split by various classifications. n is number of observations, R^2^ is the coefficient of determination of a linear regression, MSE is mean-squared-error, RMSE is root-mean-squared-error, MAE is mean-absolute-error, NSE is Nash-Sutcliffe efficiency, and pbias is percent bias. {#tbl-performance}

@@

Given the variations in the period of record across the instrumented network (see Figure S2), we analyzed trends in base flow and BFI in relation to coincident trends in climate variables. Trends are classified as coincident when the direction of the climate variable trend (positive or negative) aligns with the trend observed in base flow or BFI (@tbl-coincident-trends). This analysis includes both significant and non-significant trends, which is appropriate where the influence of complex, interconnected processes may not always manifest as statistically significant patterns over limited observational periods [@ficklin2016].

@@

The utility of the streamgage networks for base-flow analyses in Arizona is limited by their design and focus. There has been an increase in the number of streamgages in the region to meet regulatory imperatives, such as the Clean Water Act. Even so, many of these gages are not suited for base-flow studies because they emphasize peak flow monitoring and lack the ability to accurately measure low-flow dynamics [@flood-control2020]. New, non-USGS streamgages are typically installed by flood control districts (e.g. the ALERT system) and are designed to track flood flows rather than base flow. New USGS streamgages have been added over the past decade to address instream flow rights and to improve the density of monitoring in the future.
